Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework

### 2017-12-14 paperIEEE.tex
review performance section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@70 db7c848b-7755-4df2-b763-8796797afa52

In 2009 we presented the initial \textsf{Equalizer} parallel rendering
framework~\cite{EMP:09}, which introduced the architecture of a generic parallel
rendering frameworks and summarized the work in parallel rendering. Since then,
an extensive Programming and User Guide provides in-depth documentation on using
and programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following related
work we assume these two publications and their references as a starting point,
and focus on the new work published since 2009.

@@

This section introduces new decomposition modes, runtime adaptations and other
optimizations to increase the rendering performance for a wide variety of
algorithms.

@@

The initial version of \textsf{Equalizer} implemented the basic sort-first
(2D), sort-last (DB),
stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
present the newly added decomposition modes and motivate their use case.
\fig{fig:compounds}
provides an overview of the new modes. We describe how to setup compound
descriptions in the framework in \cite{EMP:09}.

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to \textsf{Equalizer}
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in a multi-threaded,
multi-GPU configuration, the application needs to support running the rendering
threads asynchronously, as outlined below in \sref{sec:threading}. The output
frame
rate of the destination channel is smoothened using a {\em frame rate equalizer}
(\sref{sec:framerateEq}).

@@

Pixel compounds (\fig{fig:pixel}) decompose the destination channel by
interleaving rows or columns in image space. They are a variant of sort-first
decomposition which works well for fill-limited applications which are not
geometry bound. Source channels cannot reduce geometry load through view frustum
culling, since each source channel has almost the same frustum (only shifted by
some pixels), but applied to a reduced 2D viewport. However, the fragment load
on all source channels is very similar due to the interleaved distribution of
pixels. This functionality is transparent to \textsf{Equalizer} applications,
and the
default compositing implementation uses the OpenGL stencil buffer to blit pixels
onto the destination channel.

@@

Subpixel compounds (\fig{fig:subpixel}) are similar to pixel compounds, but they
decompose the work for a single pixel, for example when using multisampling or
depth of field. Composition typically uses accumulation and averaging of all
computed fragments for a pixel. This feature is not fully transparent to the
application, since it needs to adapt (jitter or tilt) the frustum based on the
iteration executed. Furthermore, subpixel compounds interact with idle image
refinements, e.g., they can accelerate idle anti-aliasing of a scene when the
camera and scene are not changed.

@@

{\em Equalizers} are an addition to compound trees. They modify parameters of
their respective subtree at runtime to dynamically optimize the resource usage,
by each tuning one aspect of the decomposition. Due to their nature, they are
transparent to application developers, but might have application-accessible
parameters to tune their behavior. Resource equalization is the critical
component for scalable parallel rendering, and therefore the eponym for the
\textsf{Equalizer} project name. \FIXME{modified paragraph is ok; we
have to ensure that the font differs between the Equalizer project and the
Equalizer components consistently throughout the text}

@@

Sort-first (\fig{fig:loadeq}) and sort-last load balancing is the most obvious
optimization for these parallel rendering modes. Our load equalizer is fully
transparent for application developers; that is, it uses a reactive approach
based on past rendering times. This assumes a reasonable frame-to-frame
coherency. Our implementation stores a 2D or 1D grid of the load, mapping the
load of each channel. The load is stored in normalized 2D/1D coordinates using
$\frac{time}{area}$ as the load, the contributing source channels are organized
in a binary tree, and then the algorithm balances the two branches of each level
by equalizing the integral over the area on each side.

We have implemented various tunable parameters allowing application developers
to optimize the load balancing based on the characteristics of their rendering
algorithm:

\begin{compactdesc}
\item[Damping] reduces frame-to-frame oscillations. It is a normalized scalar
  defining how much of the computed delta from the previous position is
  applied. The equal load distribution within the region of interest assumed by
  the load equalizer is in reality not equal, causing the load balancing to
  overshoot.
\item[Resistance] eliminates small deltas in the load balancing step. This might
  help the application to cache some computations since the frustum does not
  change each frame.
\item[Boundaries] define the modulo factor in pixels onto which a load split may
  fall. Some rendering algorithms produce artefacts related to the OpenGL raster
  position, e.g., screen door transparency, which can be eliminated by aligning
  the boundary to the pixel repetition. Furthermore, some rendering algorithms
  are sensitive to cache alignments, which can again be exploited by chosing the
  corresponding boundary.
\end{compactdesc}

@@

Cross-segment load balancing (\fig{fig:vieweq}) addresses the optimal resource
allocation of $n$ rendering resources to $m$ output channels (with $n\geq
m$). The view equalizer works in conjunction with load equalizer balancing the
individual output channels. It monitors the usage of shared source channels
(across outputs) and activates them to balance the rendering time of all
outputs. In \cite{EEP:11}, we provide a detailed description and evaluation of
our algorithm.

@@

The DFR equalizer (\fig{fig:dfr}) provides a functionality similar to dynamic
video resizing \cite{MBDM:97}; that is, it maintains a constant framerate by
adapting the rendering resolution of a fill-limited application. In
\textsf{Equalizer},
this works by rendering into a source channel (typically on a FBO) separate to
the destination channel, and then scaling the rendering during the transfer
(typically through an on-GPU texture) to the destination channel. The DFR
equalizer monitors the rendering performance and accordingly adapts the
resolution of the source channel and zoom factor for the source to destination
transfer. If the performance and source channel resolutions allows, this will
not only subsample, but also supersample the destination channel to reduce
aliasing artefacts.

@@

The framerate equalizer smoothens the output frame rate of a destination
channel by instructing the corresponding window to delay its buffer swap to a
minimum time between swaps. This is regularly used for time-multiplexed
decompostions, where source channels tend to drift and finish their rendering
not evenly distributed over time. This equalizer is however fully independent of
DPlex compounds, and may be used to smoothen irregular application rendering
algorithms.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
large data scenarios. To easily customize the synchronization of all threads on
a single process, \textsf{Equalizer} implements three threading modes:  Full
synchronization, draw synchronization and asynchronous. Note that the execution
between nodes is always asynchronous, for up to \textsf{latency} frames.

@@

In draw synchronization, the node thread and all render threads are synchronized
for all \textsf{frameDraw} operations; that is, \textsf{Node::frameFinish} is
executed after the last channel is done drawing. This allows the render threads
to read shared data during their draw operation, but not during
compositing. Since compositing is typically independent of the rendered data,
this is the default mode. This mode allows to overlap compositing with data
synchronization on multi-GPU machines.

@@

Virtual Reality is an important field for parallel rendering. It does however
require special attention to support it as a first-class citizen in a generic
parallel rendering framework. \textsf{Equalizer} has been used in many virtual
reality
installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
at the KAUST visualization laboratory and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed to support
these installations. All features presented were motivated by application use
cases and have been validated with their respective users.

@@

Head tracking is the minimal feature needed to support immersive
installations. \textsf{Equalizer} does support multiple, independent tracked
views
through the observer abstraction introduced in \sref{sec:observer}. Built-in
VRPN support enables the direct, application-transparent configuration of a VRPN
tracker device. Alternatively, an application can provide a $4\times 4$ tracking
matrix defining the transformation from the canvas coordinate system to the
observer. Both CAVE-like tracking with fixed projection surfaces and HMD
tracking modes are implemented.

@@

To our knowledge, all parallel rendering systems have the focal plane coincide
with the physical display surface. For better viewing comfort, we introduce a
new dynamic focus mode, where the application defines the distance of the focal
plane from the observer, based on the current view direction of the
user. Initial experiments show that this is particularly effective for objects
placed within the immersive space; that is, in front of all display segments.

@@

Traditional head tracking computes the left and right eye positions by using a
configurable interocular distance and the tracking matrix. However, human heads
are not symmetric, and by measuring individual users a more precise frustum can
be computed. \textsf{Equalizer} supports this through the optional
configuration (file or
programmatically) of individual 3D positions for the left and right eye.

@@

Applications can switch each view between mono and stereo rendering at runtime,
and can run both monoscopic and stereoscopic views concurrently
(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
does potentially involve the start and stop of resources and processes for
passive stereo or stereo-dependent task decompositions, as described in
\sref{sec:reconfig}.

>>>>

In 2009 we presented the initial \textsf{Equalizer} parallel rendering
framework~\cite{EMP:09}, which introduced the architecture of a generic parallel
rendering framework and summarized the work in parallel rendering. Since then,
an extensive Programming and User Guide provides in-depth documentation on using
and programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following related
work we assume these two publications and their references as a starting point,
and focus on the new work published since 2009.

@@

This section introduces new decomposition modes, runtime adaptations and other
optimizations to increase the rendering performance for a wide variety of
algorithms and use cases.

@@

The initial version of \textsf{Equalizer} implemented the basic sort-first (2D),
sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the
following we present the newly added decomposition modes and motivate their use
case, which bring the overall feature set way beyond the typical sort-first and
sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
modes. We describe how to setup compound descriptions in \cite{EMP:09}.

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to \textsf{Equalizer}
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in multi-threaded,
multi-GPU configurations, the application needs to support running the rendering
threads asynchronously (\sref{sec:threading}). The output frame rate of the
destination channel may be smoothened using a {\em frame rate equalizer}
(\sref{sec:framerateEq}).

@@

Pixel compounds (\fig{fig:pixel}) decompose the destination channel by
interleaving rows or columns in image space. They are a variant of sort-first
decomposition which works well for fill-limited applications which are not
geometry bound, for example direct volume rendering. Source channels cannot
reduce geometry load through view frustum culling, since each source channel has
almost the same frustum. However, the fragment load on all source channels is
reduced linearly and well load-balanced due to the interleaved distribution of
pixels. This functionality is transparent to \textsf{Equalizer} applications,
and the default compositing uses the stencil buffer to blit pixels onto the
destination channel.

@@

Subpixel compounds (\fig{fig:subpixel}) are similar to pixel compounds, but they
decompose the work for a single pixel, for example when using FSAA or depth of
field. Composition typically uses accumulation and averaging of all computed
fragments for a pixel. This feature is not fully transparent to the application,
since it needs to adapt (jitter or tilt) the frustum based on the iteration
executed. Furthermore, subpixel compounds interact with idle image refinements,
e.g., they can accelerate idle anti-aliasing of a scene when the camera and
scene are not changed.

@@

{\em Equalizers} are an addition to compound trees. They modify parameters of
their respective subtree at runtime to dynamically optimize the resource usage,
by each tuning one aspect of the decomposition. Due to their nature, they are
transparent to application developers, but might have application-accessible
parameters to tune their behavior. Resource equalization is the critical
component for scalable parallel rendering, and therefore the eponym for the
\textsf{Equalizer} project name.

@@

Sort-first (\fig{fig:loadeq}) and sort-last load balancing is the most obvious
optimization for these parallel rendering modes. Our load equalizer is fully
transparent for application developers; that is, it uses a reactive approach
based on past rendering times. This assumes a reasonable frame-to-frame
coherency. Our implementation stores a 2D or 1D grid of the load, mapping the
load of each channel. The load is stored in normalized 2D/1D coordinates using
$\frac{time}{area}$ as the load, the contributing source channels are organized
in a binary tree, and then the algorithm balances the two branches of each level
by equalizing the integral over the area on each side. We have implemented
various tunable parameters allowing application developers to optimize the load
balancing based on the characteristics of their rendering algorithm:

\begin{compactdesc}
\item[Damping] reduces frame-to-frame oscillations. The equal load distribution
within the region of interest assumed by the load equalizer is in reality not
equal, causing the load balancing to overshoot. Damping is a normalized scalar
defining how much of the computed delta from the previous position is applied to
the new split.
\item[Resistance] eliminates small deltas in the load balancing step. This might
help the application to cache visibility computations since the frustum does not
change each frame.
\item[Boundaries] define the modulo factor in pixels onto which a load split may
fall. Some rendering algorithms produce artefacts related to the OpenGL raster
position, e.g., screen door transparency, which can be eliminated by aligning
the boundary to the pixel repetition. Furthermore, some rendering algorithms are
sensitive to cache alignments, which can again be exploited by chosing the
corresponding boundary.
\end{compactdesc}

@@

Cross-segment load balancing (\fig{fig:vieweq}) addresses the optimal resource
allocation of $n$ rendering resources to $m$ output channels (with $n\geq m$). A
view equalizer works in conjunction with load equalizers balancing the
individual output channels. It monitors the usage of shared source channels
across outputs and activates them to balance the rendering time of all outputs.
In \cite{EEP:11}, we provide a detailed description and evaluation of our
algorithm.

@@

The DFR equalizer (\fig{fig:dfr}) provides a functionality similar to dynamic
video resizing \cite{MBDM:97}, that is, it maintains a constant framerate by
adapting the rendering resolution of a fill-limited application. In
\textsf{Equalizer}, this works by rendering into a source channel (typically on
a FBO) separate to the destination channel, and then scaling the rendering
during the transfer (typically through an on-GPU texture) to the destination
channel. The DFR equalizer monitors the rendering performance and accordingly
adapts the resolution of the source channel and zoom factor for the source to
destination transfer. If the performance and source channel resolutions allows,
this will not only subsample, but also supersample the destination channel to
reduce aliasing artefacts.

@@

The framerate equalizer smoothens the output frame rate of a destination channel
by instructing the corresponding window to delay its buffer swap to a minimum
time between swaps. This is regularly used for time-multiplexed decompostions,
where source channels tend to drift and finish their rendering unevenly
distributed over time. This equalizer is however fully independent of DPlex
compounds, and may be used to smoothen irregular application rendering
algorithms.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
many complex applications and large data scenarios. To easily customize the
synchronization of all threads on a single process, \textsf{Equalizer}
implements three threading modes:  Full synchronization, draw synchronization
and asynchronous. Note that the execution between node processes is always
asynchronous, for up to \textsf{latency} frames.

@@

In draw synchronization, the node thread and all render threads are synchronized
for all \textsf{frameDraw} operations; that is, \textsf{Node::frameFinish} is
executed after the last channel is done drawing. This allows the render threads
to read shared data during their draw operation, but not during compositing.
Since compositing is typically independent of the rendered data, this is the
default mode. This mode allows to overlap compositing with rendering and data
synchronization on multi-GPU machines.

@@

Virtual Reality is an important field for parallel rendering. It does however
require special attention to support it as a first-class citizen in a generic
parallel rendering framework. \textsf{Equalizer} has been used in many virtual
reality installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution
C6 CAVE at the KAUST visualization laboratory, and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed to support
these installations. All features presented were motivated by application use
cases and have been validated with their respective users.

@@

Head tracking is the minimal feature needed to support immersive installations.
\textsf{Equalizer} does support multiple, independent tracked views through the
observer abstraction (\sref{sec:observer}). Built-in VRPN support enables the
direct, application-transparent configuration of a VRPN tracker device.
Alternatively, an application can provide a $4\times 4$ tracking matrix defining
the transformation from the canvas coordinate system to the observer. Both
CAVE-like tracking with fixed projection surfaces and HMD tracking with moving
displays are implemented.

@@

To our knowledge, all parallel rendering systems have the focal plane coincide
with the physical display surface. For better viewing comfort, we introduce a
new dynamic focus mode, where the application defines the distance of the focal
plane from the observer, based on the current view direction of the
user. Initial experiments show that this is particularly effective for objects
placed within the immersive space, that is, in front of the physical displays.

@@

Traditional head tracking computes the left and right eye positions by using a
configurable interocular distance and the tracking matrix. However, human heads
are not symmetric, and by measuring individual users a more precise frustum can
be computed. \textsf{Equalizer} supports this through the optional configuration
(file or programmatically) of individual 3D translations for all eyes relative
to the tracking matrix.

@@

Applications can switch each view between mono and stereo rendering at runtime,
and can run both monoscopic and stereoscopic views concurrently
(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
does potentially involve the start and stop of resources and processes for
passive stereo or stereo-dependent task decompositions (\sref{sec:reconfig}).
